Inferring Lensing Mass in the Universe \\ from Photometric Catalog Data

### 2012-07-02 doc/paper-lcr.tex
started re-write of theory

%Time delay distances[✓], and the mass sheet degeneracy[✓] - how kappa is degenerate in
%strong lens models (and magnifications).

Strong lensing occurs when a massive object and background source are almost perfectly
aligned along a line of sight. The light from the background source is deflected by the
lens galaxy; this deflection allows multiple images of the background source to form
at stationary points of the time delay function. For an isolated lens, the time delay
function can be calculated from calculated from 
\be \label{eq:T} 
\Delta t(\bmath{\theta},\bmath{\beta}) = \frac {1}{c} \frac{D_{\rm d} D_{\rm s}}{D_{\rm ds}} (1+z_{\rm d})\, \phi(\bmath{\theta},\bmath{\beta}),
\ee
where $\bmath{\theta}$ is the observed source position, $\bmath{\beta}$ is the 
unlensed source position, $z_{\rm d}$ is the redshift of the lens, $\phi(\bmath{\theta},\bmath{\beta})$ is
the Fermat potential, and $D_{\rm d}$, $D_{\rm s}$, and $D_{\rm ds}$
are, respectively, the angular diameter distances from observer to the lens, from observer
to the source, and from the lens to the source. The Fermat potential is given by
defined as
\be \label{eq:FP}
\phi(\bmath{\theta},\bmath{\beta})\equiv \left[\frac{(\bmath{\theta}-\bmath{\beta})^2}{2}-\psi(\bmath{\theta}) \right], 
\ee
where $\psi(\bmath{\theta})$ is the lens potential, derived from the projected dimensionless
surface mass density, $\kappa(\bmath{\theta})$, or the physical projected surface mass density,
$\Sigma(D_{\rm d} \bmath{\theta})$, by 
\be \label{eq:psikappa}
\kappa(\bmath{\theta})= \frac {\Sigma(D_{\rm d} \bmath{\theta})} {\Sigma_{\rm cr}}=\frac{1}{2}\nabla^2\psi(\bmath{\theta}),
\ee
where 
\be \label{eq:sigcrit} 
\Sigma_{\rm cr} \equiv \frac{c^2 D_{\rm s}}{4 \pi G D_{\rm d} D_{\rm ds}}.
\ee

Strong lenses sit in our real Universe -- they are not truely isolated;
there is some mass along the line of sight coming from intervening structures.
These structures may be physically associated with the lens e.g. other members
of a group or cluster, or they may be the result of chance alignments at any redshift. 
Whilst it is rare for three galaxies to line up well enough for both of the background sources
to be strongly lensed \citep{gavazzi2008,collett2012}, the shear scale of dark matter halos
makes a partial allignment -- with the additional structure acting as a weak lens -- 
a near certainty, indeed \citet{valewhite2003} and \citet{hilbert2007} find that there
are no empty lightcones in a realistic universe. These additional structures induce an external shear
which can be recovered in the mass-modelling of the lens, but they can also introduce a convergence that
is undetectable from the image positions, shapes and relative fluxes; {\it the mass sheet-degeneracy} \citep[see e.g.][for details]{Falco1985,Schneider2006}. Similar to Eq. \ref{eq:psikappa} the convergence from an isolated mass sheet is by the ratio of the projected surface mass density devided by the critical density at the redshift of the mass sheet.
\flag{but is this the kappa that enters the lens modelling?}
In the case of several mass sheets at redshifts $\{z_{\mathrm{p}}\}$, with surface mass densities $\{\Sigma_{\mathrm{p}}\}$, the effective convergence is given by the \flag{weighted?} sum:
\be \label{eq:kappasum}
\kappax=\frac{4\pi G}{c^2}\sum_{\mathrm{p}}\frac{\Sigma_{\mathrm{p}}D_{\mathrm{p}} D_{{\rm ps}}}{D_{\rm s}}.
\ee

If the source brightness changes, the observed brightness in each image does not change
simultaneously - there is a time delay, $\Delta t$, between the brightness change in
each image. Despite the invariance of the image positions and fluxes, under the addition
of an external convergence the relative Fermat potential between the images is {\it not}
invariant, and so the time delay changes. As such it is necessary to include $\kappax$
in our lens modelling, if cosmological parameters are to be estimated accurately and
precisely \citep{suyu2010} from observed time delays.

The time-delay distance is defined as
\be \label{eq:dt}
\Dt \equiv (1+\zd) D_{\rm d} D_{\rm s}/D_{\rm ds},
\ee
with which Equation \ref{eq:Tsimp} can be rephrased as
\be
\Delta t(\bmath{\theta},\bmath{\beta})  =  \frac{\Dt}{c},
\ee 
Since the time delays are proportional to $1/ (1+ \kappax)$, if there is any external convergence, $\kappax$, that is not included in the lens
modeling, then the time delay distance -- inferred assuming $\kappax = 0$ -- will be
$1-\kappax$ less than the true value of $\Dt$.
\be 
\label{eq:MassSheet:H0bias}
\Dt^{\rm{true}}=(1-\kappax) \Dt^{{\kappax = 0}}.
\ee
We can overcome this degeneracy if we have additional knowledge of $\kappax$.

For mass-sheets that are physically associated with the lens galaxy, the mass sheet
 affects the stellar dynamics of the
lens galaxy, so dynamical observations can break the internal
mass-sheet degeneracy by providing an additional estimate of the lens galaxy's mass
\citep[e.g.,][]{GroginNarayan96a, GroginNarayan96b, TonryFranx99, KoopmansTreu02,
 TreuKoopmans02, BarnabeKoopmans07, suyu2010}.

In this work our aim is to quantify the lensing effect of mass-sheets that are physically associated with the lens galaxy -- mass-sheets
lying infront or behind the lens -- and the uncertainties induced by reconstructing
those mass-sheets using simple prescriptions applied to potential astronomical observables.
We also aim to quantify how uncertainties in the
estimations of $\kappax$ propogate into uncertainties on the cosmological parametres
which can be inferred from measurements of $\Dt$.

@@

With this catalogue of halos, positions, redshifts etc. it is possible to attempt
a reconstruction of the convergernce induced by every halo near a line of sight.
Cosmological dark matter simulations have shown that dark matter halos are well
approximated by NFW halos \citep{NFW1997}, with a density profile given by
\be\label{eq:rhonfw}
\rho(r) = 
\frac{\rho_0}{\left(r/r_{s}\right)
\left(1+r/r_{s}\right)^{2}},
\ee
where $r_{s}$ is a characteristic radius of the cluster, representing where
the density slope transitions from $r^{-1}$ to $r^{-3}$. This radius is related
to the virial radius of the halo by $r_{s} = r_{200}/c$, where $r_{200}$ is the radius at 
which 
\be
\rho(r_{200}) = 200\rho_{{\mathrm{crit}}}(z), \text{ where } \rho_{\mathrm{crit}}(z) \equiv \frac{3H^{2}(z)}{8\pi G}
\ee
 and $c$ is the concentration parameter, which can be estimated from the halo's mass,
using a mass--concentration relation. Typically more massive halos are more concentrated,
but there is some scatter; we use the relation of \citet{neto2007} to estimate $c$
from the mass enclosed within $r_{200}$, which we denote as $M_{200}$. $\rho_0$ is calculated 
by the critical density and concentration parameter:
\be
\rho_0 = \rho_{\mathrm{crit}}(z\frac{200}{3}\frac{c^{3}}{\ln(1+c)-c/(1+c)}.
\ee

>>>>

%Time delay distances, and the mass sheet degeneracy - how kappa is degenerate in
%strong lens models (and magnifications).

Define: convergence.

Where does it crop up? where is it important?

The convergence from an isolated mass sheet is the ratio of the projected surface mass density devided by the critical density at the redshift of the mass sheet.
\be
\kappa(\bmath{\theta})= \frac {\Sigma(D_{\rm d} \bmath{\theta})} {\Sigma_{\rm cr}}
\ee
where 
\be \label{eq:sigcrit} 
\Sigma_{\rm cr} \equiv \frac{c^2 D_{\rm s}}{4 \pi G D_{\rm d} D_{\rm ds}}.
\ee

Strong lensing occurs when a massive object and background source are almost perfectly
aligned along a line of sight. The light from the background source is deflected by the
lens galaxy; this deflection allows multiple images of the background source to form
at stationary points of the time delay function. For an isolated lens, the time delay
function can be calculated from
\be \label{eq:T} 
\Delta t(\bmath{\theta},\bmath{\beta}) = \frac {1}{c} \frac{D_{\rm d} D_{\rm s}}{D_{\rm ds}} (1+z_{\rm d})\, \phi(\bmath{\theta},\bmath{\beta}),
\ee
where $\bmath{\theta}$ is the observed source position, $\bmath{\beta}$ is the 
unlensed source position, $z_{\rm d}$ is the redshift of the lens, $\phi(\bmath{\theta},\bmath{\beta})$ is
the Fermat potential, and $D_{\rm d}$, $D_{\rm s}$, and $D_{\rm ds}$
are, respectively, the angular diameter distances from observer to the lens, from observer
to the source, and from the lens to the source. The Fermat potential is given by
\be \label{eq:FP}
\phi(\bmath{\theta},\bmath{\beta})\equiv \left[\frac{(\bmath{\theta}-\bmath{\beta})^2}{2}-\psi(\bmath{\theta}) \right], 
\ee
where $\psi(\bmath{\theta})$ is the lens potential, derived from the projected dimensionless
surface mass density, $\kappa(\bmath{\theta})$, or the physical projected surface mass density,
$\Sigma(D_{\rm d} \bmath{\theta})$, by 
\be \label{eq:psikappa}
\kappa(\bmath{\theta})= \frac {\Sigma(D_{\rm d} \bmath{\theta})} {\Sigma_{\rm cr}}=\frac{1}{2}\nabla^2\psi(\bmath{\theta}).
\ee

Strong lenses sit in our real Universe -- they are not truely isolated;
there is some mass along the line of sight coming from intervening structures.
These structures may be physically associated with the lens e.g. other members
of a group or cluster, or they may be the result of chance alignments at any redshift. 
Whilst it is rare for three galaxies to line up well enough for both of the background sources
to be strongly lensed \citep{gavazzi2008,collett2012}, the shear scale of dark matter halos
makes a partial allignment -- with the additional structure acting as a weak lens -- 
a near certainty, indeed \citet{valewhite2003} and \citet{hilbert2007} find that there
are no empty lightcones in a realistic universe. These additional structures induce an external shear
which can be recovered in the mass-modelling of the lens, but they can also introduce a convergence that
is undetectable from the image positions, shapes and relative fluxes; {\it the mass sheet-degeneracy} \citep[see e.g.][for details]{Falco1985,Schneider2006}.
In the case of several mass sheets at redshifts $\{z_{\mathrm{p}}\}$, with surface mass densities $\{\Sigma_{\mathrm{p}}\}$, the effective convergence is given by the \flag{weighted?} sum:
\be \label{eq:kappasum}
\kappax=\frac{4\pi G}{c^2}\sum_{\mathrm{p}}\frac{\Sigma_{\mathrm{p}}D_{\mathrm{p}} D_{{\rm ps}}}{D_{\rm s}}.
\ee

If the source brightness changes, the observed brightness in each image does not change
simultaneously - there is a time delay, $\Delta t$, between the brightness change in
each image. Despite the invariance of the image positions and fluxes, under the addition
of an external convergence the relative Fermat potential between the images is {\it not}
invariant, and so the time delay changes. As such it is necessary to include $\kappax$
in our lens modelling, if cosmological parameters are to be estimated accurately and
precisely from observed time delays \citep{suyu2010}.

The time-delay distance is defined as
\be \label{eq:dt}
\Dt \equiv (1+\zd) D_{\rm d} D_{\rm s}/D_{\rm ds},
\ee
with which Equation \ref{eq:T} can be rephrased as
\be
\Delta t(\bmath{\theta},\bmath{\beta})  =  \frac{\Dt}{c} \, \phi(\bmath{\theta},\bmath{\beta})    ,
\ee 
Since the time delays are proportional to $1/ (1+ \kappax)$, if there is any external convergence, that is not included in the lens
modeling, then the time delay distance -- inferred assuming $\kappax = 0$ -- will be
$(1-\kappax)$ less than the true value of $\Dt$.
\be 
\label{eq:MassSheet:H0bias}
\Dt^{\rm{true}}=(1-\kappax) \Dt^{{\kappax = 0}}.
\ee
We can overcome this degeneracy if we have additional knowledge of $\kappax$.

For mass-sheets that are physically associated with the lens galaxy, the mass sheet
 affects the stellar dynamics of the
lens galaxy, so dynamical observations can break the internal
mass-sheet degeneracy by providing an additional estimate of the lens galaxy's mass
\citep[e.g.,][]{citations}.

In this work our aim is to quantify the lensing effect of mass-sheets that are not physically associated with the lens galaxy -- mass-sheets
lying infront or behind the lens -- and the uncertainties induced by reconstructing
those mass-sheets using simple prescriptions applied to potential astronomical observables.
We also aim to quantify how uncertainties in the
estimations of $\kappax$ propogate into uncertainties on the cosmological parametres
which can be inferred from measurements of $\Dt$.

@@

With this catalogue of halos, positions, redshifts etc. it is possible to attempt
a reconstruction of the convergernce induced by every halo near a line of sight.
Cosmological dark matter simulations have shown that dark matter halos are well
approximated by NFW profiles \citep{NFW1997}, with a density profile given by
\be\label{eq:rhonfw}
\rho(r) = 
\frac{\rho_0}{\left(r/r_{s}\right)
\left(1+r/r_{s}\right)^{2}},
\ee
where $r_{s}$ is a characteristic radius of the cluster, representing where
the density slope transitions from $r^{-1}$ to $r^{-3}$. This radius is related
to the virial radius of the halo by $r_{s}~=~r_{200}/c$, where $r_{200}$ is the radius at 
which 
\be
\rho(r_{200}) = 200\rho_{{\mathrm{crit}}}(z), \text{ where } \rho_{\mathrm{crit}}(z) \equiv \frac{3H^{2}(z)}{8\pi G}
\ee
 and $c$ is the concentration parameter, which can be estimated from the halo's mass,
using a mass--concentration relation. Typically more massive halos are more concentrated,
but there is some scatter; we use the relation of \citet{neto2007} to estimate $c$
from the mass enclosed within $r_{200}$, which we denote as $M_{200}$. $\rho_0$ is calculated 
by the critical density and concentration parameter:
\be
\rho_0 = \rho_{\mathrm{crit}}(z)\frac{200}{3}\frac{c^{3}}{\ln(1+c)-c/(1+c)}.
\ee
